An Event-based Architecture for Multi-population Optimization Algorithms

### 2020-01-20 draft.tex
Adressing comments

\begin{abstract} 
Splitting a population into multiple instances is a technique that has been used extensively in
recent years to improve the performance of nature-inspired optimization
algorithms. Work on several populations can be done in parallel, and
also asynchronously, which can be leveraged to create scalable
implementations based on, among other methods, distributed, multi-threaded, parallel, and cloud-based computing. 
% I have rephrased all this. Please check it out - JJ
The design of these cloud-native, distributed, multi-population
algorithms is not a trivial task. Coming from monolithic
(single-instance) solutions, adaptations at several levels, from the
algorithmic to the functional, must be made to leverage
the scalability, elasticity, fault-tolerance, reproducibility, and
cost-effectiveness of cloud systems while, at the same time,
conserving the intended functionality. Instead of an evolutive
approach, in this paper we propose to create from scratch a
cloud-native optimization framework that is able to include multiple
algorithms and use few parameters to work. This solution goes beyond
current state of the art via a framework that is able to support
different algorithms at the same time, work asynchronously, and also
be easily deployable to any cloud platform. We evaluate the
performance and scalability of this solution, together with the effect
other design parameters had on the 
performance, in particular, the number and the size of populations
with respect to problem size. The architecture and the implemented platform is an excellent
alternative for locally or in the cloud, thus proving that cloud-native
bioinspired algorithms perform better in their "natural" environment than other
kinds of algorithms, and setting a new baseline for scaling and
performance in the cloud.
% Still a big longish. We need another iteration to reduce it - JJ
\end{abstract}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{kafkEOsmall}
    \caption{Example of a parametric plot ($\sin (x), \cos(x), x$)} % This caption? - JJ
    \label{fig:kafkEO}
\end{figure}

@@

The primary data element used throughout the system is the subpopulation, and at a higher
level, we can view the system as a continuous stream of subpopulations
flowing
% We talk about "multi-population" elsewhere, including the
% abstract. We should unify terminology - JJ
from one process to the other. By "continuous stream," we mean that ideally,
everything must happen asynchronously without components needing to wait for
others. In this kind of system, this streaming functionality is normally
implemented by using a message queue system. For instance, if one process needs
to communicate a subpopulation to another, it must pack the sub-population into
a message, and then send the message asynchronously % I don't know if
                                % asynchronously applies here. It's
                                % simply sent - JJ
to the queuing system; this
means that after the process pushes the message, it continues its execution
without waiting for a result. On the other end, the recipient subscribes to the
message queue by defining an event handler method that is going to be triggered when a
message is pushed to the message queue. The two components at each end of the queue are
Producers that push messages to the queue and Consumers that pull them. This
pattern is an essential component of a reactive architecture because it is
highly scalable; one of the reasons for this are stateless functions. We can
define stateless functions as methods that do not have side effects, meaning
they do not need to read, keep, or modify data outside of the method. If
Consumers are implemented using stateless functions, then there is no difference
between having one or many copies of the same function pulling work from the queue all at the
same time, because no side effects are going to result from the
concurrency.
% This is important and we should probably mention that in the
% introduction - JJ 
Based on these general principles we define next the proposed model.

\subsection{Event Driven Model} 
\label{edm}
Based on a reactive architecture we proposed in a previous work
\cite{guervos2018introducing}, we now describe the general architecture shown in
Figure \ref{fig:kafkEO}. Again, we can explain the model using the analogy
of Producers and Consumers of messages. First, we can see that
there are two queues, one labeled Input, and the other one Output. In the diagram, 
push operations on a queue are represented by solid arrows connecting to the left side
of the box, and pull or pop operations as solid arrows leaving from the right side.
The algorithm starts with the Setup process pulling a configuration message from
the experiements queue, this is not shown in the diagram. The configuration messsage 
includes all the parameters needed to execute the algorithm, among other things,
the number of subpopulations, the number of individuals, and the number of iterations
of the algorithm. We give more details about the configuration data structure
in Section []. Once the configuration is read, we can follow the path of 
messagess as follows:

\begin{enumerate}
\item In this step, the specified number of subpopulations is created acording to the 
parameters found in the configuration structure. The population at this moment is just
static data, including each individual inside. Each subpopulation includes a metadata
section where its algorithm and execution parameters are specified. For instance, 
for a GA, the mutation rate, type of crossover operator, and other values are indicated.
Each subpopulation is then pushed to the Input queue, so they can be consumed 
by Stateless functions responsible for the execution of the search.  % Why the caps in Stateless? - JJ

\item One or more Stateless functions are constantly pulling subpopulation messages
from the Input queue. Taking the population data and metadata as parameters, these 
functions are responsible for the actual execution of the optimization algorithm. 
They take the current state of the population and run the algorithm for a certain 
number of iterations. Once they finish the execution, the population state
is packed again along with aditional metadata about the execution of the algorithm.
The resulting populations is now pushed to the Output queue. Once finished, another 
subpopulation is pulled from the queue.

\item The Controller process is responsible for keeping track of the progress of 
the search. It pulls current subpopulations from the Output queue, inspects the metadata 
and if an optimal solution has been found or the maximum number of iteration has been
reached it signals the stop of the execution. Otherwise, it passes the stream of messages 
to a migration process, where subpopulations are mixed with each other. New populations 
are generated from this migration, and they are again pushed to the Input queue to continue 
in a loop. The Controller is also responsible of logging the metadata recived with the
messages.  
\end{enumerate}

This reactive architecture has the following advantages:\begin{itemize}
\item 
An important aspect of proposal is the decoupling of the population and 
the population-based algorithm. It is common that in a classic island-based algorithm
each island is executed in a separate processing node, i.e. in a virtual machine, cpu,
or thread. In this case we can have just one processing node, or many nodes, each 
running a different search strategy, or using its own parameters. 
\item 
Also, the reactive controller gives designers more control over the multi-population algorithm.
In this process, designers can dynamicly change the number of populations, population parameters,
and migration details on-the-fly.
\item 
Another advantege is that algorithm designers have many options for implementing this simple architecture. 
The same basic components can be implmented as a single multithreaded program, or 
as a highly scalable serverless cloud application. Most modern languages include 
constructs for asynchronous programming using queues or channels, for a multi-threaded
execution.
\end{itemize}

>>>>

\begin{abstract} 
Splitting a population into multiple instances is a technique that has been used extensively in
recent years to improve the performance of nature-inspired optimization
algorithms. Work on several populations can be done in parallel, and
also asynchronously, which can be leveraged to create scalable
implementations based on, among other methods, distributed, multi-threaded, parallel, and cloud-based computing. 
% I have rephrased all this. Please check it out - JJ
The design of these cloud-native, distributed, multi-population
algorithms is not a trivial task. Coming from monolithic
(single-instance) solutions, adaptations at several levels, from the
algorithmic to the functional, must be made to leverage
the scalability, elasticity, fault-tolerance, reproducibility, and
cost-effectiveness of cloud systems while, at the same time,
conserving the intended functionality. Instead of an evolutive
approach, in this paper we propose to create from scratch a
cloud-native optimization framework that is able to include multiple
algorithms and use few parameters to work. This solution goes beyond
current state of the art via a framework that is able to support
different algorithms at the same time, work asynchronously, and also
be easily deployable to any cloud platform. We evaluate the
performance and scalability of this solution, together with the effect
other design parameters had on the 
performance, in particular, the number and the size of populations
with respect to problem size. The architecture and the implemented platform is an excellent
alternative for running locally or in the cloud, thus proving that cloud-native
bioinspired algorithms perform better in their "natural" environment than other
kinds of algorithms, and setting a new baseline for scaling and
performance in the cloud.
% Still a big longish. We need another iteration to reduce it - JJ
\end{abstract}

@@

@@

The primary data element used throughout the system is the subpopulation, and at a higher
level, we can view the system as a continuous stream of subpopulations
flowing
% We talk about "multi-population" elsewhere, including the
% abstract. We should unify terminology - JJ
% "multi-population" is the whole thing, a subpopulation is a member of that.
% "A multi-population has many subpopulations or should we call them just populations?"
% I will add a comment about this in the Multi-population section - Mario
from one process to the other. By "continuous stream," we mean that ideally,
everything must happen asynchronously without components needing to wait for
others. In this kind of system, this streaming functionality is normally
implemented by using a message queue system. For instance, if one process needs
to communicate a subpopulation to another, it must pack the sub-population into
a message, and then send the message asynchronously % I don't know if
                                % asynchronously applies here. It's
                                % simply sent - JJ 
                                % Meaning it does not wait for a respone, it continues doing other stuff - Mario 
to the queuing system; this
means that after the process pushes the message, it continues its execution
without waiting for a result. On the other end, the recipient subscribes to the
message queue by defining an event handler method that is going to be triggered when a
message is pushed to the message queue. The two components at each end of the queue are
Producers that push messages to the queue and Consumers that pull them. This
pattern is an essential component of a reactive architecture because it is
highly scalable; one of the reasons for this are stateless functions. We can
define stateless functions as methods that do not have side effects, meaning
they do not need to read, keep, or modify data outside of the method. If
Consumers are implemented using stateless functions, then there is no difference
between having one or many copies of the same function pulling work from the queue all at the
same time, because no side effects are going to result from the
concurrency.
% This is important and we should probably mention that in the
% introduction - JJ 
Based on these general principles we define next the proposed model.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{kafkEOsmall}
    \caption{The proposed general architecture, 
     showing each process in a swimlane, message-dataflow,
     message queues, and high-level tasks for each process.} % This caption? - JJ
    \label{fig:kafkEO}
\end{figure}

\subsection{Event Driven Model} 
\label{edm}
Based on a reactive architecture we proposed in a previous work
\cite{guervos2018introducing}, we now describe the general architecture shown in
Figure \ref{fig:kafkEO}. Again, we can explain the model using the analogy
of Producers and Consumers of messages. First, we can see that
there are two queues, one labeled Input, and the other one Output. In the diagram, 
push operations on a queue are represented by solid arrows connecting to the left side
of the box, and pull or pop operations as solid arrows leaving from the right side.
Also, the architecture has at least four processes indicated in the diagram as
swimlanes: First, the Setup process, responsible for the reading a configuration
file, and creating the initial subpopulations. Second, a Controller process,
responsible for the migration between subpopulations, and keeping track of the
iterations of the algorithm. The Message Queue process runs the Input and Output
queues. Finally, there is at least one Stateless Function process responsible
for running the isolated algorithms. In the example, two processes PSO and GA,
are shown.

The algorithm starts with the Setup process pulling a configuration message from
the experiements queue, this is not shown in the diagram. The configuration messsage 
includes all the parameters needed to execute the algorithm, among other things,
the number of subpopulations, the number of individuals, and the number of iterations
of the algorithm. We give more details about the configuration data structure
in Section []. Once the configuration is read, we can follow the path of 
messagess as follows:

\begin{enumerate}
\item In this step, the specified number of subpopulations is created acording to the 
parameters found in the configuration structure. The population at this moment is just
static data, including each individual inside. Each subpopulation includes a metadata
section where its algorithm and execution parameters are specified. For instance, 
for a GA, the mutation rate, type of crossover operator, and other values are indicated.

\item Each subpopulation is then pushed to the Input queue, so they can be consumed 
by Stateless functions responsible for the execution of the search.  % Why the caps in Stateless? - JJ  
% It refers to the swimlines in the diagram, maybe I should point it out  - Mario
% Added a comment above. - Mario

\item One or more Stateless functions are constantly pulling subpopulation messages
from the Input queue. Taking the population data and metadata as parameters, these 
functions are responsible for the actual execution of the optimization algorithm. 
They take the current state of the population and run the algorithm for a certain 
number of iterations.

\item Once they finish the execution, the population state
is packed again along with aditional metadata about the execution of the algorithm.
The resulting populations are now pushed to the Output queue. Once finished, another 
subpopulation is pulled from the queue.

\item The Controller process is responsible for keeping track of the progress of 
the search. It pulls current subpopulations from the Output queue, inspects the metadata 
and if an optimal solution has been found or the maximum number of iteration has been
reached it signals the stop of the execution.

\item Otherwise, it passes the stream of messages 
to a migration process, where subpopulations are mixed with each other. New populations 
are generated from this migration, and they are again pushed to the Input queue to continue 
in a loop. The Controller is also responsible of logging the metadata recived with the
messages.  
\end{enumerate}

This reactive architecture has the following advantages:\begin{itemize}
\item 
An important aspect of proposal is the decoupling of the population and 
the population-based algorithm. It is common that in a classic island-based algorithm
each island is executed in a separate processing node, i.e. in a virtual machine, cpu,
or thread. In this case we can have just one processing node, or many nodes, each 
running a different search strategy, or using its own parameters. 
\item 
Also, the reactive controller gives designers more control over the multi-population algorithm.
In this process, designers can dynamically change the number of populations, population parameters,
and migration details on-the-fly.
\item 
Another advantege is that algorithm designers have many options for implementing this simple architecture. 
The same basic components can be implmented as a single multithreaded program, or 
as a highly scalable serverless cloud application. Most modern languages include 
constructs for asynchronous programming using queues or channels, for a multi-threaded
execution.
\end{itemize}
